[Artifact Presentation] Damegender: \\ Writing and Comparing Gender Detection Tools

### 2020-05-15 paper2sattose.tex

In Section 2, we explain Damegender as solution to the
problem. Section 3 explains that we are using Open Datasets extracted
from the states as base of truth. Section 4, presents a feature
comparison with other tools.  Section 5, reproduces accuracies and
confusion matrix with a scientific dataset. The section 6 is about how
we are using Machine Learning in Damegender. The section 7 is about
conclusion, limitations and futher research.

@@

@@

\begin{table*}[ht]
\footnotesize
\begin{tabular}[]{lcccccc}
\hline
Service / Tool $->$ & Gender API & gender-guesser & genderize.io & NameAPI & NamSor & damegender\tabularnewline
\hline
Database size & 431M & 45K & 114M & 1M & 4G & 57K \tabularnewline
Regular data updates & Yes & no & No & Yes & Yes & Yes\tabularnewline
Unstructured full name strings & Yes & No & No & Yes & No & Yes\tabularnewline
Surnames & Yes & No & No & Yes & Yes & Yes\tabularnewline
Non-Latin alphabets & Partial & No & Partial & Yes & Yes & No\tabularnewline
Implicit geo-localization & Yes & No & No & Yes & Yes & No\tabularnewline
Exists locale & Yes & Yes & Yes & Yes & Yes & Yes\tabularnewline
Assingment type & P & B & P & P & P & P \tabularnewline
Free parameters & T,P & G & P,C & T & S & T,C\tabularnewline
Prediction & No & No & No & No & No & Yes\tabularnewline
Free license & No & Yes & No & No & No & Yes\tabularnewline
REST API & Yes & No & Yes & Yes & Yes & Planned\tabularnewline
Limits number of requests & Yes (200) & \infinity & Yes & Yes & Yes & \infinity \tabularnewline
\hline
\end{tabular}
\caption{Comparison of the different features that gender guesser software services and tools offer. Assignment type = \{P: Probabilistic; B: Binary\}. Free Parameters = \{T: total\_names; P: probability; C: count; G: gender; T: trust; S: scale \}. }
\label{table:comparison}
\end{table*}

\grex{Estaria bien contar con ``the monthly subscription cost (calculated for 100,000 requests/month)'' en la tabla, porque daría más info muy relevante.}
\davidam{Si te da tiempo por mi adelante}

@@

Santamaria and Mihaljevic~\cite{10.7717/peerj-cs.156} offers a framework to classify gender tools.
The features observed in this framework are: (i) database size (as of January 2018), (ii) if there are regular data updates, (iii) if they handle unstructured full name strings, (iv) if they handle surnames, (v) if they handle non-Latin alphabets, (vi) if implicit geolocalization is available, (vii) if locale exists, (viii) the type of assignment, (ix) if free parameters are possible, (x) if they offer prediction, (xi) if the tool is released under an open source license, (xii) if they offer an API, (xiii) the amount of monthly free requests, and (xiv) the monthly subscription cost (calculated for 100,000 requests/month)).

@@

Santamaria and Mihaljevic~\cite{10.7717/peerj-cs.156} explains different ways to determine gender from a name by humans and offers 7,000 names applying these methods. 
In their dataset, gender is classified as male, female or unknown. 
We have used this dataset, not considering the unknow variable, for our experiments.
The results can be seen in Table~\ref{table:DifferentAccuraciesMeasures}.
As can be observed, Genderapi and Genderize are obtaining the best results, although all solutions reach results better than 0.8, except for \texttt{Gender Guesser}.

\grex{Deberiamos explicar que es cada una de las metricas. Y tambien por que tenemos 1.0 de recall en todos los casos.}
\davidam{Esto es un poco largo de explicar. Está explicado de una manera abreviada más arriba. Si te ves con energías para explicarlo, está explicado en https://easychair.org/publications/preprint/vthL}

%\begin{table*}[]{@{}lllllll@{}}
\begin{table}
\footnotesize
\begin{tabular}[]{lcccc}
  \hline
  API & Acc & Prec & F1 & Recall\tabularnewline
\hline
Genderapi & 0.969 & 0.972 & 0.964 & 1.0\tabularnewline
Genderize & 0.927 & 0.976 & 0.966 & 1.0\tabularnewline
Damegender (SVC)\footnote{SVC is the acronym of Support Vector Classification, the Machine Learning algorithm that Damegender was using with this results} & 0.879 & 0.972 & 0.972 & 1.0\tabularnewline
Namsor & 0.867 & 0.973 & 0.924 & 1.0\tabularnewline
Nameapi & 0.830 & 0.974 & 0.905 & 1.0\tabularnewline
Gender Guesser & 0.774 & 0.985 & 0.872 & 1.0\tabularnewline
\hline
\end{tabular}
\caption{Different accuracies measures}
\label{table:DifferentAccuraciesMeasures}
\end{table}

@@

\begin{table}
\footnotesize
\begin{tabular}[]{lrrrr}
  \hline
  APIs          & gender & male & female & undef \tabularnewline
\hline
%%\endhead
Genderapi         & male    & 3589 & 155  &  67 \tabularnewline
                  & female  & 211  & 1734 &  23 \tabularnewline
Damegender       & male    & 3663 & 147  &   0 \tabularnewline
(SVC)\footnotemark[1] & female  & 551  & 1497 &   0 \tabularnewline
Genderguesser     & male    & 3326 &  139 & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Namsor            & male    & 3325 & 139  & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Genderize         & male    & 3157 & 242  & 412 \tabularnewline
                  & female  & 75   & 1742 & 151 \tabularnewline
Nameapi           & male    & 2627 & 674  & 507 \tabularnewline
                  & female  & 667  & 1061 & 240 \tabularnewline 
\hline
\end{tabular}
\caption{Confusion matrix tables by APIs}
\label{table:ConfusionMatrixTables}
\end{table}

In Table~\ref{table:ApisAndErrors} we can observe the different measures for errors in the APIs. The most visible conclusion is an high index of errors in \texttt{Nameapi} and \texttt{Namsor} and a low index of errors in \texttt{GenderApi} and \texttt{damegender}.

\begin{table*}
\footnotesize
\begin{tabular}[]{lrrrr}
\hline
API & error & error w/o na & na coded & error gender bias\tabularnewline
\hline
%\endhead
Damegender (SVC)\footnotemark[1] & 0.121 & 0.121 & 0.0 & -0.07\tabularnewline
GenderApi & 0.167 & 0.167 & 0.0 & -0.167\tabularnewline
Gender Guesser & 0.225 & 0.027 & 0.204 & 0.003\tabularnewline
Genderize & 0.276 & 0.261 & 0.0204 & -0.0084 \tabularnewline 
Namsor & 0.332 & 0.262 & 0.095 & 0.01 \tabularnewline
Nameapi & 0.361 & 0.267 & 0.129 & 0.001 \tabularnewline
\hline
\end{tabular}
\caption{APIs and Errors}
\label{table:ApisAndErrors}
\end{table*}

@@

We have developed a script infofeatures.py with our datasets to analyze data about features.
The datasets used in this experiment was retrieved from official datasets from national statistical
institutions in Spain, Uruguay, United Kingdom, USA. The features used are: first letter, last letter, [a-z], vocals, consonants, first letter, first letter vocal, last letter vocal, last letter consonant, last letter a. 
The choosing of features was verified with Principal Component Analysis.
The most relevant results for the different datasets are offered in Table~\ref{table:InfoFeatures}.

\begin{table*}
\footnotesize
\begin{tabular}[]{lrrrrrrr}
  \hline
Dataset & a & last a & last o & last consonant & last vocal & 1st consonant & 1st vocal  \tabularnewline
  \hline
  %\endhead
 Uruguay (F) \footnote{F is for females and M is for males} &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177  \tabularnewline
 Uruguay (M) \footnotemark[3] &    0.643 &         0.249 &         0.062 &                 0.766 &             0.234 &                  0.771 &              0.228  \tabularnewline
 Spain (F) \footnotemark[3]  &    0.922 &         0.588 &          0.03 &                 0.271 &             0.728 &                  0.772 &              0.228  \tabularnewline
 Spain (M) \footnotemark[3]  &    0.818 &          0.03 &         0.268 &                 0.569 &              0.43 &                  0.763 &              0.236  \tabularnewline
 UK (F) \footnotemark[3]     &    0.825 &         0.374 &         0.013 &                 0.322 &             0.674 &                  0.765 &              0.235  \tabularnewline
 UK (M) \footnotemark[3]     &    0.716 &         0.036 &         0.039 &                  0.78 &             0.218 &                  0.799 &                0.2  \tabularnewline
 USA (F) \footnotemark[3]    &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177  \tabularnewline
 USA (M) \footnotemark[3]    &    0.643 &          0.02 &         0.061 &                 0.765 &             0.234 &                   0.84 &              0.159  \tabularnewline
 Canada (F) \footnotemark[3] &    0.659 &         0.189 &         0.005 &                 0.591 &             0.408 &                  0.838 &              0.16  \tabularnewline
 Canada (M) \footnotemark[3] &    0.752 &          0.22 &         0.025 &                  0.54 &             0.456 &                  0.818 &              0.181  \tabularnewline

\hline
\end{tabular}
\caption{Informative Features in Different Countries}
\label{table:InfoFeatures}
\end{table*}

The countries where the main language is Spanish (Uruguay + Spain) and English (USA + United Kingdom + Australia) where is having very similar variation with the features chosen between males and females with these datasets (remember that these datasets are being extracted from official statistics provided by the states). 
In Canada, a country French centric has different rules with this features.

The letter a is varying 0.2 from males to females in (USA and Uruguay) and 0.1 from males to females (United Kingdom, Australia and Spain).
The last letter a is varying 0.5 from males to females in (Australia, Spain) around 0.4 in (USA, United Kingdom) and 0.2 in
Uruguay.
The last letter o from females to males is varying 0.2 in (Spain, Australia) and is equal in (Uruguay, USA, United Kingdom).
For the last letter consonant all countries is giving the result that is for males, with results from 0.2 to 0.5: Uruguay and USA (0.5), United Kingdom (0.4), Australia and Spain (0.3). 
So last letter vocal is reverse than last letter consonant. 
First letter consonant or first letter vocal is a non significant feature due to so similar results in English and Spanish.
The relevant metrics with the different algorithms are shown in Table~\ref{table:MLAccuracies}.

\begin{table}
\footnotesize
\begin{tabular}[]{lcccc}
  \hline
  ML Algorithm & Acc & Prec & F1 & Recall \tabularnewline
  \hline
  %\endhead
 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0  \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0  \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0  \tabularnewline
 Stoch. Gradient Distrib.    &    0.709 &     0.943 &   0.815 &    1.0  \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0  \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0  \tabularnewline

>>>>

In Section 2 we explain \texttt{damegender} as solution to the
problem of gender guessing. Section 3 introduces the Open Datasets used as the \emph{golden truth}.
Section 4 presents a feature comparison with other gender guession tools and services.
Section 5 reports on accuracies and offers a confusion matrix using a scientific dataset.
Section 6 is about how we use Machine Learning in \texttt{damegender}.
Section 7 discusses limitations and futher research, and concludes the paper.

@@

\begin{table*}[ht]
\footnotesize
\begin{tabular}[]{lcccccc}
\hline
Service / Tool $->$ & Gender API & gender-guesser & genderize.io & NameAPI & NamSor & damegender\tabularnewline
\hline
Database size & 431M & 45K & 114M & 1M & 4G & 57K \tabularnewline
Regular data updates & Yes & no & No & Yes & Yes & Yes\tabularnewline
Unstructured full name strings & Yes & No & No & Yes & No & Yes\tabularnewline
Surnames & Yes & No & No & Yes & Yes & Yes\tabularnewline
Non-Latin alphabets & Partial & No & Partial & Yes & Yes & No\tabularnewline
Implicit geo-localization & Yes & No & No & Yes & Yes & No\tabularnewline
Exists locale & Yes & Yes & Yes & Yes & Yes & Yes\tabularnewline
Assingment type & P & B & P & P & P & P \tabularnewline
Free parameters & T,P & G & P,C & T & S & T,C\tabularnewline
Prediction & No & No & No & No & No & Yes\tabularnewline
Free license & No & Yes & No & No & No & Yes\tabularnewline
REST API & Yes & No & Yes & Yes & Yes & Planned\tabularnewline
Limits number of requests & Yes (200) & \infinity & Yes & Yes & Yes & \infinity \tabularnewline
Subscription (100K requests/month)	 & 79 & 0 & 7 & 150 & 80 & 0 \tabularnewline
\hline
\end{tabular}
\caption{Comparison of the different features that gender guesser software services and tools offer. Assignment type = \{P: Probabilistic; B: Binary\}. Free Parameters = \{T: total\_names; P: probability; C: count; G: gender; T: trust; S: scale \}. The subscription price is given in euro.}
\label{table:comparison}
\end{table*}

@@

@@

Santamaria and Mihaljevic~\cite{10.7717/peerj-cs.156} offer a framework to classify gender tools.
The features observed in this framework are: (i) database size (as of January 2018), (ii) if there are regular data updates, (iii) if they handle unstructured full name strings, (iv) if they handle surnames, (v) if they handle non-Latin alphabets, (vi) if implicit geolocalization is available, (vii) if locale exists, (viii) the type of assignment, (ix) if free parameters are possible, (x) if they offer prediction, (xi) if the tool is released under an open source license, (xii) if they offer an API, (xiii) the amount of monthly free requests, and (xiv) the monthly subscription cost (calculated for 100,000 requests/month)).

@@

Santamaria and Mihaljevic~\cite{10.7717/peerj-cs.156} explain different ways to determine gender from a name by humans and offer 7,000 names applying these methods. 
In their dataset, gender is classified as male, female or unknown. 
We have used this dataset, not considering the unknow variable, for our experiments.
The results, using common information retrieval metrics, can be seen in Table~\ref{table:DifferentAccuraciesMeasures}.
Accuracy is the ratio of correctly predicted observation to the total observations.
It should be noted that \emph{Accuracy} can be a misleading metric for imbalanced data sets, such as the ones that we usually have in software development projects.
So, for a sample with 85 negative and 15 positive values, classifying all values as negative in this case gives a score of accuracy of 0.85.
In those cases, it is better to report other measures, such as the balanced accuracy (bACC), which normalizes true positive and true negative predictions by the number of positive and negative samples~\cite{mower2005prep}.
Precision is the fraction of relevant instances among the retrieved instances.
Recall is the fraction of the total amount of relevant instances that were actually retrieved.
In our case, we have left no name out, so recall is 1 for all tools.
Precision and recall are sometimes used together in the F1 Score (or f-measure) to provide a single measurement for a system.
As can be observed, Genderapi and Genderize obtain the best results -- although all solutions are close and reach results better than 0.8 for accuracy, except for \texttt{Gender Guesser}.

%\grex{Deberiamos explicar que es cada una de las metricas. Y tambien por que tenemos 1.0 de recall en todos los casos.}
%\davidam{Esto es un poco largo de explicar. Está explicado de una manera abreviada más arriba. Si te ves con energías para explicarlo, está explicado en https://easychair.org/publications/preprint/vthL}

%\begin{table*}[]{@{}lllllll@{}}
\begin{table}[t]
\footnotesize
\begin{tabular}[]{lcccc}
  \hline
  API & Acc & Prec & F1 & Recall\tabularnewline
\hline
Genderapi & 0.969 & 0.972 & 0.964 & 1.0\tabularnewline
Genderize & 0.927 & 0.976 & 0.966 & 1.0\tabularnewline
Damegender (SVC)\footnote{SVC is the acronym of Support Vector Classification, the Machine Learning algorithm that Damegender was using with this results} & 0.879 & 0.972 & 0.972 & 1.0\tabularnewline
Namsor & 0.867 & 0.973 & 0.924 & 1.0\tabularnewline
Nameapi & 0.830 & 0.974 & 0.905 & 1.0\tabularnewline
Gender Guesser & 0.774 & 0.985 & 0.872 & 1.0\tabularnewline
\hline
\end{tabular}
\caption{Different accuracies measures}
\label{table:DifferentAccuraciesMeasures}
\end{table}

@@

\begin{table}[t]
\footnotesize
\begin{tabular}[]{lrrrr}
  \hline
  APIs          & gender & male & female & undef \tabularnewline
\hline
%%\endhead
Genderapi         & male    & 3589 & 155  &  67 \tabularnewline
                  & female  & 211  & 1734 &  23 \tabularnewline
Damegender       & male    & 3663 & 147  &   0 \tabularnewline
(SVC)\footnotemark[1] & female  & 551  & 1497 &   0 \tabularnewline
Genderguesser     & male    & 3326 &  139 & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Namsor            & male    & 3325 & 139  & 346 \tabularnewline
                  & female  & 78   & 1686 & 204 \tabularnewline
Genderize         & male    & 3157 & 242  & 412 \tabularnewline
                  & female  & 75   & 1742 & 151 \tabularnewline
Nameapi           & male    & 2627 & 674  & 507 \tabularnewline
                  & female  & 667  & 1061 & 240 \tabularnewline 
\hline
\end{tabular}
\caption{Confusion matrix tables by APIs}
\label{table:ConfusionMatrixTables}
\end{table}

In Table~\ref{table:ApisAndErrors} we observe the different measures for errors in the APIs.
Error coded defines if the true is different than the guessed one. 
Error coded without na defines if the true is different than the guessed one, but without undefined results.
Error gender bias allows to understand if the error is bigger for guessing males than females or viceversa.
The weighted error defines if the true value is different than the guessed one, but giving a weight to the guessed as undefined.
The most relevant information is a high index of errors for \texttt{Nameapi} and \texttt{Namsor},
while \texttt{GenderApi} and \texttt{damegender} have a low index of errors.

\begin{table*}
\footnotesize
\center
\begin{tabular}[]{lrrrr}
\hline
API & error & error w/o na & na coded & error gender bias\tabularnewline
\hline
%\endhead
Damegender (SVC)\footnotemark[1] & 0.121 & 0.121 & 0.0 & -0.07\tabularnewline
GenderApi & 0.167 & 0.167 & 0.0 & -0.167\tabularnewline
Gender Guesser & 0.225 & 0.027 & 0.204 & 0.003\tabularnewline
Genderize & 0.276 & 0.261 & 0.0204 & -0.0084 \tabularnewline 
Namsor & 0.332 & 0.262 & 0.095 & 0.01 \tabularnewline
Nameapi & 0.361 & 0.267 & 0.129 & 0.001 \tabularnewline
\hline
\end{tabular}
\caption{APIs and Errors}
\label{table:ApisAndErrors}
\end{table*}

@@

We have developed some experimental functionality that allows to analyze our database according to some features using machine learning algorithms.
To test our approach, we have selected some features of the names, such as a being the first letter, a (or o) being the last letter, contains the letter a, first letter is a vocal, last letter is a vocal, last letter is a consonant, or last letter is a. 
The selection of the features was verified with Principal Component Analysis.
The datasets used in this experiment were the ones from the Spain National Institute of Statistics and the Natural Language
ToolKit corpus of (English) names.
The most relevant results for the different datasets used are offered in Table~\ref{table:InfoFeatures}.

\begin{table*}
\footnotesize
\center
\begin{tabular}[]{lccccccc}
  \hline
Dataset & contains a & last is a & last is o & last is consonant & last is vocal & 1st is consonant & 1st is vocal  \tabularnewline
  \hline
  %\endhead
 Uruguay (F) \footnote{F is for females and M is for males} &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177  \tabularnewline
 Uruguay (M) \footnotemark[3] &    0.643 &         0.249 &         0.062 &                 0.766 &             0.234 &                  0.771 &              0.228  \tabularnewline
 Spain (F) \footnotemark[3]  &    0.922 &         0.588 &          0.03 &                 0.271 &             0.728 &                  0.772 &              0.228  \tabularnewline
 Spain (M) \footnotemark[3]  &    0.818 &          0.03 &         0.268 &                 0.569 &              0.43 &                  0.763 &              0.236  \tabularnewline
 UK (F) \footnotemark[3]     &    0.825 &         0.374 &         0.013 &                 0.322 &             0.674 &                  0.765 &              0.235  \tabularnewline
 UK (M) \footnotemark[3]     &    0.716 &         0.036 &         0.039 &                  0.78 &             0.218 &                  0.799 &                0.2  \tabularnewline
 USA (F) \footnotemark[3]    &    0.816 &         0.456 &         0.007 &                 0.287 &             0.712 &                  0.823 &              0.177  \tabularnewline
 USA (M) \footnotemark[3]    &    0.643 &          0.02 &         0.061 &                 0.765 &             0.234 &                   0.84 &              0.159  \tabularnewline
 Canada (F) \footnotemark[3] &    0.659 &         0.189 &         0.005 &                 0.591 &             0.408 &                  0.838 &              0.16  \tabularnewline
 Canada (M) \footnotemark[3] &    0.752 &          0.22 &         0.025 &                  0.54 &             0.456 &                  0.818 &              0.181  \tabularnewline
\hline
\end{tabular}
\caption{Informative Features in Different Countries. F stands for females, and M for males.}
\vspace{0.3cm}
\label{table:InfoFeatures}
\end{table*}

As expected, countries that share language offer similar results, i.e., the variation of the chosen featuers between males and females is comparable.
This is the case for Uruguay and Spain (Spanish, and USA, UK and Australia (English).
In Canada, a country that has an ample French-speaking community these features show a different trend.

For instance, the letter a varies 0.2 from males to females in (USA and Uruguay) and 0.1 from males to females (United Kingdom, Australia and Spain).
The last letter a varies 0.5 from males to females in (Australia, Spain), around 0.4 in (USA, United Kingdom) and 0.2 in
Uruguay.
The last letter o from females to males varies 0.2 in (Spain, Australia) and is equal in Uruguay, USA, United Kingdom.
For the last letter consonant all countries give as a result that males do have it more frequently, with results that range from 0.3 to 0.5: Uruguay and USA (0.5), United Kingdom (0.4), Australia and Spain (0.3). 
So, last letter vocal is reverse than last letter consonant. 
First letter consonant or first letter vocal is a non-significant feature due to offering similar results in English and Spanish.

The relevant metrics with the different algorithms are shown in Table~\ref{table:MLAccuracies}.

\begin{table}
\footnotesize
\center
\begin{tabular}[]{lcccc}
  \hline
  ML Algorithm & Acc & Prec & F1 & Recall \tabularnewline
  \hline
  %\endhead
 Support Vector Machines             &    0.879 &     0.972 &   0.972 &    1.0  \tabularnewline
 Random Forest                       &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 NLTK (Bayes)                        &    0.862 &     0.902 &   0.902 &    1.0  \tabularnewline
 Multinomial Navie Bayes             &    0.782 &     0.791 &   0.791 &    1.0  \tabularnewline
 Tree                                &    0.764 &     0.821 &   0.796 &    1.0  \tabularnewline
 Stoch. Gradient Distrib.    &    0.709 &     0.943 &   0.815 &    1.0  \tabularnewline
 Gaussian Naive Bayes                &    0.709 &     0.968 &   0.887 &    1.0  \tabularnewline
 Bernoulli Naive Bayes               &    0.699 &     0.965 &   0.816 &    1.0  \tabularnewline
